Fail Faster

### 2025-03-16 paper/paper.tex
another pass

Property-based testing (PBT) is a testing technique where random data
produced by a \textit{generator} is fed
to a system under test, which is then checked against a set of properties 
it is expected to satisfy. Recent work on understanding how PBT is used in 
practice has determined that practitioners often run PBT with very short time 
budgets---typically 50 milliseconds to 30 seconds. Given this practical 
constraint, it is imperative that inputs be generated as fast as possible; 
in other words, generators must be treated as performance-sensitive code.
A high-leverage way to improve the performance of generators is to optimize 
\textit{generator libraries}: if
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance experts.

However, the performance of 
existing generator libraries falls short of what is possible due to two major 
sources of inefficiency. The first of these is \textit{abstraction}: while 
flexible and expressive, generator libraries rely on complex abstractions
and indirections that are difficult for compilers to fully simplify, resulting
 in additional function calls and closure allocations.
Our second insight involves the process of 
selecting random values (``random sampling''), which, depending on one's choice of \rand and
the number of samples needed, can constitute an unexpectedly high proportion
of a generator's runtime. This raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To answer this, we propose a two-pronged approach to designing generator libraries 
that preserves their expressiveness while eliminating unnecessary overhead.
First, we propose a method of optimizing
generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
long been used across domains as a technique of eliminating the overhead 
of idiomatic functional DSLs \todo{CITE!!!}.
Using staging, it is possible to eliminate the overhead of many common 
generator abstractions. For example,
each call to monadic bind in \bq constructs and then immediately executes a 
closure at runtime; meanwhile, in \name, bind resolves at compile time,
allowing the compiler to inline away the closures. Secondly, we demonstrate the
impact of different choices of \rand on generator performance 
by conducting a controlled comparison. Together, these insights provide a
recipe for improving even state-of-the-art generator libraries.

@@

To demonstrate the effectiveness of our technique, we apply it to \bq,
which is the fastest property-based testing library in OCaml according to our tests.
%\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
Our version of this library, \name, is a drop-in replacement for \bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name 
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons between generators, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values. Therefore, differences in performance are directly 
attributable to \name's speedup.

To evaluate \name's effectiveness, 
implement a series of workloads in each generator library, and demonstrate that 
\name runs faster and allocates less than \bq. Further, we show that this performance benefit
translates to bug-finding ability by running these workloads in Etna, a PBT platform that 
simulates real-world usage by injecting bugs into a system and measuring how quickly different 
generators can detect them.
Finally, to demonstrate the generality of our technique, we implement \name in Scala
and benchmark the same generators in the Scala port of the library---thereby highlighting 
its portability
to other strict functional languages and reinforcing its potential for widespread adoption.

In summary, we demonstrate that, across languages, monadic PBT generator DSLs
have a significant performance overhead, and present a cross-language technique
for eliminating that overhead while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that showcases our optimization technique while maintaining strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}

>>>>

Property-based testing (PBT) is a technique in which a \textit{generator} produces random inputs to a system under test, which is then evaluated against a set of properties the system is expected to satisfy.
Recent studies on PBT usage have shown that practitioners often run tests with very short time budgets---typically between 50 milliseconds and 30 seconds. Given this constraint, generating inputs as quickly as possible is crucial for maximizing the odds of finding a bug in the limited time available. In other words, generators must be treated as performance-sensitive code.
A high-leverage way to improve the performance of generators is to optimize 
\textit{generator libraries}: if
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance engineers.

However, the performance of 
existing generator libraries falls short of what is possible. This is due to two major 
sources of inefficiency. First, the same high-level design that gives generator libraries their flexibility and expressiveness also introduces overhead:
layered abstractions and indirections hinder the compiler's ability to optimize computations, shifting more work to runtime.
Second, the process of 
selecting random values (``random sampling'') can constitute an unexpectedly high proportion
of a generator's runtime, depending on one's choice of \rand and
the number of samples needed. This raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To answer this, we propose a two-pronged approach for designing generator libraries 
that preserves their expressiveness while eliminating unnecessary overhead.
First, we describe a method of optimizing
generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
long been used to eliminating overhead 
from idiomatic functional DSLs \todo{CITE!!!}.
Using staging, it is possible to eliminate the overhead of many common 
generator abstractions. For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
Secondly, we demonstrate the
impact of different choices of \rand on generator performance 
by conducting a controlled comparison. Together, these insights provide a
recipe for improving even state-of-the-art generator libraries.

@@

To demonstrate the effectiveness of our technique, we apply it to \bq, the fastest property-based testing library in OCaml according to our tests.
%\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
Our staged version of this library,\\ \name, is a direct drop-in replacement for \bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \\\name 
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values. Therefore, differences in performance are directly 
attributable to \name's speedup.

To evaluate \name's effectiveness, we
implement a series of workloads in each generator library, and show that 
\name runs faster and allocates less than \\\bq. Further, we show that this performance benefit
translates to bug-finding ability by running these workloads in Etna, a platform that 
simulates real-world PBT usage by injecting bugs into a system and measuring how quickly various 
generators detect them.
Finally, to demonstrate the generality of our technique, we implement \name in Scala
and benchmark the same generators in the Scala port of the library---thereby highlighting 
its portability
to other strict functional languages and reinforcing its potential for widespread adoption.

In summary, we establish that monadic PBT generator DSLs
have a significant performance overhead across languages, and present a cross-language technique
for eliminating that overhead while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that showcases our optimization technique while maintaining strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}
